$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-05-06 src/s_ar1.Rmd
added sensitivity plots

>>>>

# Show performance as a function of hyperparameters
```{r}
caretGrid <- expand.grid(interaction.depth=c(3), n.trees = 50,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
caretGrid1 <- expand.grid(interaction.depth=c(5), n.trees = 50,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
caretGrid2 <- expand.grid(interaction.depth=c(3), n.trees = 100,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
caretGrid3 <- expand.grid(interaction.depth=c(5), n.trees = 100,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
caretGrid4 <- expand.grid(interaction.depth=c(7), n.trees = 100,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
caretGrid5 <- expand.grid(interaction.depth=c(7), n.trees = 200,
                   shrinkage=c(0.1),
                   n.minobsinnode=10)
lgrids <- list(caretGrid, caretGrid1, caretGrid2, caretGrid3, caretGrid4, caretGrid5)
f_replicate_hyper <- function(lgrids){
  temp <- f_generate_lower_var_four(1/3, 0.3, 1, 1000)
  a_array <- array(dim=c(1000, 4, 1))
  a_array[,,1] <- temp
  laccuracy <- vector(length = length(lgrids))
  for(i in seq_along(lgrids))
    laccuracy[i] <- r_star(a_array, caret_default=lgrids[[i]])
  return(laccuracy)
}

nreplicates <- 200
r_star_vals <- matrix(nrow = nreplicates, ncol=length(lgrids))
for(i in 1:nreplicates){
  if(i%%100==0)
    print(i)

  temp <- f_replicate_hyper(lgrids)
  r_star_vals[i, ] <- temp
}

saveRDS(r_star_vals, "../data/ar1_sensitivity.rds")

b_sum <-
  r_star_vals %>% 
  as.data.frame() %>% 
  mutate(iterate=seq(1, nreplicates, 1)) %>%
  melt(id.vars="iterate") %>% 
  mutate(below=if_else(value < 1, 1, 0)) %>% 
  group_by(variable) %>% 
  summarise(below_per=round(mean(below)*100, 1),
            lower=quantile(value, 0.25),
            upper=quantile(value, 0.75),
            middle=median(value)) %>%
  mutate(below_per=paste0(below_per, "%")) %>% 
  mutate(x_pos=seq(1, 6, 1)) %>% 
  mutate(y_pos=0.95)

g <- r_star_vals %>% 
  as.data.frame() %>% 
  mutate(iterate=seq(1, nreplicates, 1)) %>% 
  melt(id.vars="iterate") %>% 
  mutate(below=if_else(value < 1, 0, 1)) %>% 
  left_join(b_sum, by = "variable") %>% 
  mutate(variable=case_when(variable=="V1"~"int.depth=3\nn.trees=50",
                            variable=="V2"~"int.depth=5\nn.trees=50",
                            variable=="V3"~"int.depth=3\nn.trees=100",
                            variable=="V4"~"int.depth=5\nn.trees=100",
                            variable=="V5"~"int.depth=7\nn.trees=100",
                            variable=="V6"~"int.depth=7\nn.trees=200")) %>% 
  ggplot(aes(x=as.factor(variable), y=value, colour=as.factor(below))) +
  geom_jitter() +
  xlab("Hyperparameters") +
  ylab("R*") +
  geom_hline(yintercept = 1, linetype=2) +
  annotate("text", x = b_sum$x_pos, y = b_sum$y_pos, label = b_sum$below_per, size=5) +
  annotate("pointrange", x = 1:6, ymin=b_sum$lower, ymax=b_sum$upper, y = b_sum$middle, colour="black", size=1) +
  coord_flip() +
  scale_color_grey() +
  theme(axis.text = element_text(colour="black", size=12),
        axis.title = element_text(colour="black", size=14),
        legend.position = "none")
ggsave("../output/ar1_sensitivity.pdf", g, width = 12, height = 8)
```

### 2020-05-06 src/s_ovarian_prostate.Rmd
added sensitivity plots

mon <- monitor(fit)
saveRDS(mon, "../data/ovarian_long_mon.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star <- r_star_dist(full_data)
saveRDS(r_star, "../data/ovarian_long_r_star.rds")
```

@@

mon <- monitor(fit)
saveRDS(mon, "../data/ovarian_short_mon.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star <- r_star_dist(full_data)
saveRDS(r_star, "../data/ovarian_short_r_star.rds")
```

@@

g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
  bind_rows(tibble(long=mon_longer$Rhat, short=mon_shorter$Rhat, metric="B. Rhat")) %>% 
  bind_rows(tibble(long=mon_longer$Bulk_ESS, short=mon_shorter$Bulk_ESS, metric="C. Bulk ESS")) %>% 
  bind_rows(tibble(long=mon_longer$Tail_ESS, short=mon_shorter$Tail_ESS, metric="D. Tail ESS")) %>% 
  melt(id.vars="metric") %>% 
  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.6) +
  facet_wrap(~metric, scales="free") +
  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
  theme(strip.text.x = element_text(size = 10, colour="black"),
        text = element_text(size=12, colour="black"),
        axis.text = element_text(size=10, colour="black"),
        legend.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank()) +
  ylab("Count")
ggsave("../output/ovarian.pdf", g, width = 12, height = 8)
```

# Prostate

```{r}
prostate <- read.csv("../data/prostate.csv", header=FALSE)

@@

mon <- monitor(fit)
saveRDS(mon, "../data/prostate_long_mon.rds")
full_data <- rstan::extract(fit, permuted=F)

@@

mon <- monitor(fit)
saveRDS(mon, "../data/prostate_short_mon.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star <- r_star_dist(full_data[, , seq(1, 18106, 5)])
saveRDS(r_star, "../data/prostate_short_r_star.rds")
```

@@

g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
  bind_rows(tibble(long=mon_longer$Rhat, short=mon_shorter$Rhat, metric="B. Rhat")) %>% 
  bind_rows(tibble(long=mon_longer$Bulk_ESS, short=mon_shorter$Bulk_ESS, metric="C. Bulk ESS")) %>% 
  bind_rows(tibble(long=mon_longer$Tail_ESS, short=mon_shorter$Tail_ESS, metric="D. Tail ESS")) %>% 
  melt(id.vars="metric") %>% 
  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.6) +
  facet_wrap(~metric, scales="free") +
  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
  theme(strip.text.x = element_text(size = 10, colour="black"),
        text = element_text(size=12, colour="black"),
        axis.text = element_text(size=10, colour="black"),
        legend.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank()) +
  ylab("Count") +
  scale_x_continuous(limits=my_breaks)
ggsave("../output/prostate.pdf", g, width = 12, height = 8)
```

>>>>

x <- rstan::extract(fit, permuted=F)
mon <- monitor_extra(split_data(x))
saveRDS(mon, "../data/ovarian_long_mon.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star <- r_star_dist(full_data)
saveRDS(r_star, "../data/ovarian_long_r_star.rds")
```

@@

x <- rstan::extract(fit, permuted=F)
mon <- monitor_extra(split_data(x))
saveRDS(mon, "../data/ovarian_short_mon.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star <- r_star_dist(full_data)
saveRDS(r_star, "../data/ovarian_short_r_star.rds")
```

@@

g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. split-Rhat")) %>% 
  bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk ESS")) %>% 
  bind_rows(tibble(long=mon_longer$tailseff, short=mon_shorter$tailseff, metric="D. Tail ESS")) %>% 
  melt(id.vars="metric") %>% 
  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. split-Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.6) +
  facet_wrap(~metric, scales="free") +
  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
  theme(strip.text.x = element_text(size = 10, colour="black"),
        text = element_text(size=12, colour="black"),
        axis.text = element_text(size=10, colour="black"),
        legend.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank()) +
  ylab("Count")
ggsave("../output/ovarian.pdf", g, width = 12, height = 8)
```

# Prostate
```{r}
prostate <- read.csv("../data/prostate.csv", header=FALSE)

@@

x <- rstan::extract(fit, permuted=F)
mon <- monitor_extra(split_data(x))
saveRDS(mon, "../data/prostate_long_mon.rds")
full_data <- rstan::extract(fit, permuted=F)

@@

x <- rstan::extract(fit, permuted=F)
mon <- monitor_extra(split_data(x))
saveRDS(mon, "../data/prostate_short_mon.rds")
full_data <- rstan::extract(fit, permuted=F)
r_star <- r_star_dist(full_data[, , seq(1, 18106, 5)])
saveRDS(r_star, "../data/prostate_short_r_star.rds")
```

@@

g <- tibble(long=r_star_longer, short=r_star_shorter, metric="A. R*") %>% 
  bind_rows(tibble(long=mon_longer$zsRhat, short=mon_shorter$zsRhat, metric="B. split-Rhat")) %>% 
  bind_rows(tibble(long=mon_longer$zsseff, short=mon_shorter$zsseff, metric="C. Bulk ESS")) %>% 
  bind_rows(tibble(long=mon_longer$tailseff, short=mon_shorter$tailseff, metric="D. Tail ESS")) %>% 
  melt(id.vars="metric") %>% 
  mutate(metric=fct_relevel(as.factor(metric), "A. R*", "B. split-Rhat", "C. Bulk ESS", "D. Tail ESS")) %>% 
  ggplot(aes(x=value, fill=as.factor(variable))) +
  geom_histogram(position="identity", alpha=0.6) +
  facet_wrap(~metric, scales="free") +
  scale_fill_grey("MCMC info.", labels=c("4 chains: long", "16 chains: short")) +
  theme(strip.text.x = element_text(size = 10, colour="black"),
        text = element_text(size=12, colour="black"),
        axis.text = element_text(size=10, colour="black"),
        legend.text = element_text(size=12, colour="black"),
        axis.title.x = element_blank()) +
  ylab("Count") +
  scale_x_continuous(limits=my_breaks)
ggsave("../output/prostate.pdf", g, width = 12, height = 8)
```

### 2020-05-06 write_up/ms.tex
added sensitivity plots

@@

@@

\subsection{Many parameter models: ovarian and prostate analysis}
In this section, we...

>>>>

To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we perform a sensitivity analysis. In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data as were used to generate Fig. \ref{fig:ar1}. In particular, the GBM hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200). The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig. \ref{fig:ar1_sensitivity}. These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper). This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to different GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases). The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated. The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ar1_sensitivity}
\end{figure}

@@

Fig. \ref{fig:eight_schools_r_star_quantiles} also indicates a potential limitation of $R^*$: namely, that as chains are progressively thinned, those regions where chains behave most idiosyncratically can be missed, resulting in a reduction in ML classification accuracy and falsely concluding that convergence has occurred.

@@

\subsection{Many parameter models: ovarian and prostate analysis}
In this section, we analyse two Bayesian models [Aki, can you provide some description of these?]. These models have relatively many parameters (4719 for the ``ovarian'' model and 18,105 for the ``prostate'' model) and are known to be challenging to explore with MCMC. For each model, we consider two MCMC runs: one with 4 chains run with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning).

For the ovarian model, we show the results in Fig. \ref{fig:ovarian}. In Fig. \ref{fig:ovarian}A, we show the $R^*$ distributions for each model run, which show that, whereas the ``long'' model run has converged, the ``short'' one has yet to do so. Whilst it is harder to discern, this pattern is mirrored in Fig. \ref{fig:ovarian}B, since the long model has $\widehat{R}<1.01$ for all parameters [check this!], whereas the short model has a handful of parameters where this is not the case. Similarly, in Figs. \ref{fig:ovarian}C and \ref{fig:ovarian}D, which show the bulk ESS and tail ESS respectively, it is evident that, for the short model, there remain a few parameters with low effective sample sizes, whereas the long model has more consistent values for this metric.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ovarian.pdf}}
	\caption{\textbf{Ovarian example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning). In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk ESS across all parameters; and D shows tail ESS across parameters. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ovarian}
\end{figure}

For the prostate model, we show the results in Fig. \ref{fig:prostate}. Since this model has nearly four times as many parameters as the ovarian model, it was more computationally expensive to estimate $R^*$ for it. To handle this, we thinned down the parameters by a factor of 5 for the long model, recognising that, of course, this measure will make it more likely that we diagnose convergence. Despite this, both $R^*$ measures indicated that the MCMC runs had yet to converge (Fig. \ref{fig:prostate}A), which was mirrored by the other metrics considered (Figs. \ref{fig:prostate}B,C\&D).

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/prostate.pdf}}
	\caption{\textbf{Prostate example.} In all plots, we show the results from two model runs: one with 4 chains with 10,000 iterations each (1000 discarded as warm-up and thinned by a factor of 10); another with 16 chains run with 1500 iterations each (500 discarded as warm-up and no thinning). In A, $R^*$ distributions (with 1000 draws each using Algorithm \ref{alg:R_star_uncertainty}) are shown -- for the long run, these were calculated after thinning the parameters by a factor of 5; B shows rank-normalised split-$\widehat{R}$ values across all parameters; C shows bulk ESS across all parameters; and D shows tail ESS across parameters. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:prostate}
\end{figure}
